On the power of oritatami cotranscriptional folding with unary bead sequence\thanks{This work is supported in part by KAKENHI Grant-in-Aid for Challenging Research (Exploratory) No.~18K19779 and JST Program to Disseminate Tenure Tracking System No.~6F36 granted to S.~Z.~F. and S.~S.}

### 2018-12-04 tt_theorem.tex
12/3

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$. The following statements hold.
\begin{enumerate}
\item At arity $\alpha \geq 3$, if $S[k] = t$ (i.e. the tunnel that stabilizes $w[k]$ is not singular) and $S[k+1] \neq \blacksquare$, then $\#bc(C_{k-1}) > \#bc(C_k)$.
\item At arity $\alpha = 2$, if there are indices $i,j$ such that $S[i..j+1]$ is either $bbt^{(j-i-1)}b$ or $bt^\ell bt^{j-i-1-\ell}b$ for some $\ell$, then $\#bc(C_{i-1}) > \#bc(C_j)$.
\end{enumerate}
\end{theorem}

@@

\begin{lemma}
\label{TTT_exit}
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$ and arity $\alpha =2$.
%Let $w[i]$ be a bead which is stabilized at the exit of a tunnel.
If $S[h..i+1] = bt^{(i-h)}b$ for some $h<i-1$, then $\#bc(C_{i-2}) \geq \#bc(C_i)$, and hence, $\#bc(C_{h-1}) \geq \#bc(C_i)$.
If $h < i-1$, then the second inequality is strengthened as $\#bc(C_{h-1}) > \#bc(C_{i})$.
%if we assume $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$, then $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$.
\end{lemma}

@@

\begin{proof}
Let us show that $\#bc(C_{i-1}) > \#bc(C_i)$, that is, when $w[i]$ is stabilized, $w[i]$ uses at least two hands.
Let us assume $w[i]$ is able to be stabilized with using one hand. Fig.\ref{TTT_a3_w} exhibits all the three kinds of possibility of stabilized $w[i]$. Then, $w[i]$ can be also stabilized at $n_3$.

\begin{paragraph}{Case of straight}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
According to assumption, $w[i]$ uses only one hand. Therefore, any neighbors of $n_3$ are occupied according to the lemma\ref{TTT_a3_2b_lemma}. $n_3$ and the point which is stabilized $w[i]$ are free so that $n_1$ has some bond by lemma\ref{TTT_neighbor_lemma}. Accordingly, this situation is non-deterministic. Thus, $n_3$ and $n_4$ have to be occupied because of symmetry.

\item[-] Otherwise\\
Because of  $S[i] = b$, at least one of $n_1$ and $n_2$ have to be free. Let us regard that $n_1$ is free. Neighbors of $n_1$ have to be occupied and at least two neighbors of $n_{-1}$ have to be free for $n_1$ and $w[i]$. According to lemma\ref{TTT_neighbor_lemma}, $n_{-1}$ have some hand. Therefore $w[i]$ can be also stabilized $n_1$, that is, this situation is non-deterministic. Thus, one of $n_3$ and $n_4$ has to be free.
\end{itemize}

Therefore, this case is false.
\end{paragraph}

\begin{paragraph}{Case of obtuse}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
Any neighbors of $n_3$ have to be occupied but the point which is stabilized $w[i]$ is free. Thus $n_3$ has to be occupied.
\item[-] Case of $n_4$ is free\\
According to lemma\ref{TTT_a3_2b_lemma}, $n_2$ has to be occupied because $n_4$ is free. Also $n_0$ has to be occupied from lemma\ref{TTT_neighbor_lemma}. Thus, only one of $n_0, n_3$ leave some hands or both of them do not leave any hands because $w[i]$ use only one bond.\\
If $n_0$ has some hands, then $n_3$ does not have any hands so that $n_{-3}$ is occupied. Also $n_{-3}$ must not have any hands so that $n_{-2}$ is occupied and also $n_{-1}$ is occupied. Therefore any neighbors of $w[i]$ are occupied so that $w[i+1]$ cannot provide.\\
If $n_3$ has some hands, then $n_0$ does not have any hands so that $n_{-1}$ is occupied. In the same previous way, any $n_{-2}, n_{-3}$ are occupied.  Therefore any neighbors of $w[i]$ are occupied.\\
If both of $n_0, n_3$ do not have any hands, then both of $n_{-1}, n_{-3}$ are occupied. If one of $n_{-1}, n_{-3}$ has some hands, the other does not have any hands so that $n_{-2}$ is occupied. If both of $n_{-1}, n_{-3}$ do not have any hands, $n_{-2}$ has to be occupied and $n_{-2}$ has some hands.Therefore any neighbors of $w[i]$ are occupied so that $w[i+1]$ cannot provide.\\
Thus $n_3$ has to be occupied in order to yield infinite structures.
\item[-] Case of $n_2$ is free\\
Any neighbors of $n_2$ have to be occupied so that $n_0$ is occupied. Any neighbors of $n_0$ except $n_2$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_2$ has to be occupied.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\end{itemize}

Therefore, any situations contradict $S[i] = b$.

\end{paragraph}

\begin{paragraph}{Case of acute}
\begin{itemize}
\item[-] Case of $n_4$ is free\\
$n_4$ and a point which is stabilized $w[i]$ are free so that  $w[i-2]$ has some hands according to lemma\ref{TTT_neighbor_lemma}.
However, $w[i]$ can be also stabilized $n_4$ in this case. Thus, $n_4$ has to be occupied.
\item[-] Case of $n_2$ is free\\
According to lemma\ref{TTT_a3_2b_lemma}, $n_0$ has to be occupied. $n_1$ has to be also occupied because of lemma\ref{TTT_neighbor_lemma}.
We consider this case just like case of obtuse and that $n_4$ is free. Then if $w[i-2]$ binds $w[i]$, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. If $n_1$ binds $w[i]$, this case is same. Also if $n_1$ and $w[i-2]$ do not have any hand, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. Therefore, $w[i+1]$ cannot be provided.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{1}$ is occupied. Any neighbors of $n_{1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\item[-] Case of $n_1$ is free\\
Any neighbors of $n_1$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_1$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_1$ has to be occupied. 
\end{itemize}

Hence, assumption that w[i] is able to be stabilized with using one hand is false.
Therefore, when w[i] is stabilized, w[i] uses at least two hands.
\end{proof}

\begin{figure}
  \begin{center}
  \begin{tabular}{c c c}
 \begin{minipage}{0.3\hsize}
 \centering
    \begin{tikzpicture}

@@

	\foreach \theta in {60,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	 \draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
 	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(0:1)] at (60:1) {$n_{-1}$};
    \end{tikzpicture}
    \end{minipage}

@@

	\foreach \theta in {0,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	\draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
	\draw [shift=(60:1), shift=(60:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	\draw [shift=(60:1), shift=(120:1)] (-0.05,-0.05) rectangle (0.05,0.05);

 	\node[right] at (0:1) {$n_0$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(0:1)] at (60:1) {$n_{-1}$};
	\node[above, shift=(60:1)] at (60:1) {$n_{-2}$};
	\node[above, shift=(60:1)] at (120:1) {$n_{-3}$};
    \end{tikzpicture}
    \end{minipage}

@@

>>>>

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$. The following statements hold.
\begin{enumerate}
\item At arity $\alpha \geq 3$, if $S[i] = t$ (i.e. the tunnel that stabilizes $w[i]$ is not singular) and $S[i+1] \neq \blacksquare$, then $\#bc(C_{i-1}) > \#bc(C_i)$.
\item At arity $\alpha = 2$, if $n$ is the number of $bt$ in $S[i..j]$ for some $i,j$ such that $S[i] = b$ and $i < j$, then $\#bc(C_{i-2}) - \#bc(C_{j}) \geq n$.
\end{enumerate}
\end{theorem}

@@

\begin{lemma}
\label{TTT_exit}
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$ and arity $\alpha =2$.
%Let $w[i]$ be a bead which is stabilized at the exit of a tunnel.
If $S[h..i+1] = bt^{(i-h)}b$ for some $h<i-1$ and $S[i] = t_s$ or $t_o$, then $\#bc(C_{i-2}) \geq \#bc(C_i)$, and hence, $\#bc(C_{h-1}) \geq \#bc(C_i)$.
If $h < i-1$, then the second inequality is strengthened as $\#bc(C_{h-1}) > \#bc(C_{i})$.
%if we assume $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$, then $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$.
\end{lemma}

@@

\begin{figure} [tb]
  \begin{center}
  \begin{tabular}{c c c}
 \begin{minipage}{0.3\hsize}
 \centering
    \begin{tikzpicture}

@@

	\foreach \theta in {60,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	 \draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
	  \draw [shift=(180:1),shift=(120:1)] (-0.05,-0.05) rectangle (0.05,0.05);
 	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(0:1)] at (60:1) {$n_{-1}$};
	\node[above, shift=(180:1)] at (120:1) {$n_5$};
    \end{tikzpicture}
    \end{minipage}

@@

	\foreach \theta in {0,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	\draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
	\draw [shift=(60:1), shift=(60:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	\draw [shift=(60:1), shift=(120:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	  \draw [shift=(180:1),shift=(-120:1)] (-0.05,-0.05) rectangle (0.05,0.05);

 	\node[right] at (0:1) {$n_0$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(60:1)] at (0:1) {$n_{-1}$};
	\node[above, shift=(60:1)] at (60:1) {$n_{-2}$};
	\node[above, shift=(60:1)] at (120:1) {$n_{-3}$};
	\node[below, shift=(180:1)] at (-120:1) {$n_5$};
    \end{tikzpicture}
    \end{minipage}

@@

%Both $n_3$ and $n_4$ are occupied because $S[i] = t$ or $t^\times$.
%Suppose $w[i]$ were stabilized without any bond.
%Then, all neighbors of $n_3$ must be occupied as long as the bead at $n_3$ is not singular; otherwise, a bead at $n_3$ would be capable of forming a bond with $w[i]$.
%That is, $n_1$ is occupied.
%We can continue this argument in clockwise manner to fill the neighbors of $w[i]$ until a singular bead is encountered, if any.
%If needed, the same argument is applied starting from $n_4$ in counter-clockwise manner.
%Thus, all neighbors of $w[i]$ must be occupied but then $S[i+1] = \blacksquare$, contradiction. \qed

\begin{proof}
We consider cases which $w[i-1]$ is two distances away from the singular point.
Let us show that $\#bc(C_{i-1}) > \#bc(C_i)$, that is, when $w[i]$ is stabilized, $w[i]$ uses at least two hands.
Fig.\ref{TTT_a3_w} exhibits all the three kinds of possibility of stabilized $w[i]$.
Suppose $w[i]$ were able to be stabilized with using one hand in each cases.

% Then, $w[i]$ can be also stabilized at $n_3$.

\begin{paragraph}{Case of straight}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
The bead $w[i-2]$ must not have bonds because $w[i]$ is stabilized deterministically so that five neighbors of $w[i-2]$ must be occupied.
Therefore $n_5$ must be occupied.
We can continue this argument in clockwise manner to fill the neighbors of $n_3$ until $n_1$.
However, two neighbors of $n_1$ must be free because they are $n_3$ and the point of $w[i]$, contradict.

\item[-] Otherwise\\
Both $n_3$ and $n_4$ are occupied.
One of $n_1$ and $n_2$ must be free because $S[i] = b$.
We assume $n_1$ is free, then all neighbors of $n_1$ must be occupied because $n_3$ is occupied.
However, one of neighbors $n_1$ is a point which $w[i]$ is stabilized, contradict.
\end{itemize}

Therefore, this case contradict assumption.
\end{paragraph}

\begin{paragraph}{Case of obtuse}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
Any neighbors of $n_3$ have to be occupied but the point which $w[i]$ is stabilized is free. Thus $n_3$ has to be occupied.
\item[-] Case of $n_4$ is free\\
The point $n_2$ must be occupied because $n_4$ is free and $n_5$ must be occupied.
Therefore $n_0$ is occupied.
The bead $w[i]$ forms just one bond so that one of $n_0, n_3$ leave some hands or both of them do not leave any bonds.\\
If $n_0$ has some hands, then $n_3$ does not have any hands so that $n_{-3}$ is occupied. 
We can continue this argument in clockwise manner to fill the neighbors of $w[i]$.
That is, $S[i] = \blacksquare$, contradict.

If $n_3$ has some hands, then $n_0$ does not have any hands so that neighbors of $w[i]$ are occupied. In the same previous way, contradict.\\
If both of $n_0, n_3$ do not have any hands, then both of $n_{-1}, n_{-3}$ are occupied. If one of $n_{-1}, n_{-3}$ has some hands, the other does not have any hands so that $n_{-2}$ is occupied. If both of $n_{-1}, n_{-3}$ do not have any hands, $n_{-2}$ has to be occupied and $n_{-2}$ has some hands, contradict.
\item[-] Case of $n_2$ is free\\
Any neighbors of $n_2$ have to be occupied so that $n_0$ is occupied. Any neighbors of $n_0$ except $n_2$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_2$ has to be occupied.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\end{itemize}
\end{paragraph}

\begin{paragraph}{Case of acute}
\begin{itemize}
\item[-] Case of $n_4$ is free\\
The point $n_4$ and a point which is stabilized $w[i]$ are free so that  $w[i-2]$ has some bonds.
However, $w[i]$ can be also stabilized $n_4$ in this case. Thus, $n_4$ has to be occupied.
\item[-] Case of $n_2$ is free\\
The point $n_0$ has to be occupied. The point $n_1$ has to be also occupied because five neighbors of $n_0$ must be occupied.
We consider this case just like case of obtuse and that $n_4$ is free. Then if $w[i-2]$ binds $w[i]$, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. If $n_1$ binds $w[i]$, this case is same. Also if $n_1$ and $w[i-2]$ do not have any hand, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. Therefore, $w[i+1]$ cannot be provided.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{1}$ is occupied. Any neighbors of $n_{1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\item[-] Case of $n_1$ is free\\
Any neighbors of $n_1$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_1$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_1$ has to be occupied. 
\end{itemize}

Hence, $w[i]$ uses at least two bonds if $w[i-1]$ is at least two distances away the singular point.
The size of $ \hexagon^2_0 = 19$ and this number contain the singular point, its successor and its three neighbors which is it binds to. Therefore 14 times  $w[i]$ can be stabilized by forming one bond.
Therefore, a deterministic unary oritatami system at $\delta = 1$ and $\alpha = 3$ can yield only finite conformation of length is $2n + 14$.
\end{proof}
